---
output:
  html_document:
    keep_md: yes
    self_contained: true
---




```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(message = FALSE)
```

|  OS             | Build           |
|-----------------|-----------------|
| Linux x86_64 / OSX   | [![Build Status](https://travis-ci.org/jaredhuling/personalized.svg?branch=master)](https://travis-ci.org/jaredhuling/personalized)  


## Introduction to the 'personalized' package

The personalized package provides estimation methods for subgroup identification under the framework of [Chen et al (2017)](http://onlinelibrary.wiley.com/doi/10.1111/biom.12676/abstract)

Install using the **devtools** package:

```{r, eval = FALSE}
devtools::install_github("jaredhuling/personalized")
```

or by cloning and building using `R CMD INSTALL`

## Quick Usage Reference

Load the package and access help files for the main functions:
```{r, message = FALSE, warning = FALSE}
library(personalized)
```

```{r, eval = FALSE}
?fit.subgrp
?validate.subgrp
```


